### 2017-07-07 paper/lexicographic.tex
Several small changes

In context of probabilistic programs we work with probability 
spaces of the form $(\Omega,\natfilt,\probm^\sigma)$, where $\Omega$ is a 
set of runs in some $\pCFG$ and $\natfilt$ is (the smallest) sigma-algebra 
such that all the functions $\cfg{\sigma}{i}$, where $i\in \Nset_0$ and 
$\sigma$ is a scheduler, are $\natfilt$-measurable. In such a setting we can 
also consider sub-sigma-algebras $\natfilt_i$, $i\in \Nset_0$, of 
$\natfilt$, where $\natfilt_i$ is the smallest sub-sigma-algebra of 
$\natfilt$ such that all the functions $\cfg{\sigma}{j}$, $0\leq j \leq 
i$, are $\natfilt_i$-measurable. Intuitively, each set $A$ belonging to such 
an $\natfilt_i$ consists of runs whose first $i$ steps satisfy some 
property, and the probability space $(\Omega,\natfilt_i,\probm^\sigma)$ 
allows us to reason about probabilities of certain events happening in the 
first 
$i$ steps of program execution. 
%(note that since $\natfilt_i \subseteq 
%\natfilt$, we can view $\probm^\sigma$ also as a function of type 
%$\natfilt_i\rightarrow [0,1]$). 
Then, for each $A\in \natfilt_i$, the 
value $\E[\E[X|\natfilt_i]\cdot 1_A]$ represents the expected value of 
$X(\run)$ for the randomly generated run $\run$ provided that we restrict to 
runs whose
prefix of length $i$ satisfies the property given by $A$. 
The sequence $\natfilt_0,\natfilt_1,\natfilt_2,\dots$ forms a 
filtration of $\natfilt$, which we call a \emph{canonical filtration}.

@@

The strict 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{HolgerPOPL,CFNH16:prob-termination}. There is one significant difference: in these works there is an additional \emph{integrability} condition imposed on the one-dimensional process $\{X_i\}_{i=0}^{\infty}$, which requires that for each $i\geq 0$ it holds $\E[|X_i|]<\infty$ (or equivalently $\E[X_i]<\infty$, as the process is required to be non-negative). We do not impose this condition, which simplifies possible application of LexRSMs to programs with non-linear arithmetic, where, as already shown in~\cite{HolgerPOPL}, integrability of program variables is not guaranteed.
%We do not impose this condition, and indeed, it is possible to define a 3-dimensional LexRSM $\{\vec{X}_{i}\}_{i=0}^{\infty}$ such that $\E[X[3]_i]=\infty$ for some $i$ (example in \AppendixMaterial). 
The reason why integrability condition can be dropped is that it is only needed in the previous works to ensure that the conditional expectations exist and are well-defined. However, the existence of conditional expectations is also guaranteed for random variables that are real-valued non-negative, see Proposition~\ref{prop:conditional-exp-existence}. 
%This is exactly the case in both the original 1-dimensional ranking supermartingales and our generalization to LexRSMs.

>>>>

\begin{remark}
\label{rem:canfilt}
In context of probabilistic programs we work with probability 
spaces of the form $(\Omega,\natfilt,\probm^\sigma)$, where $\Omega$ is a 
set of runs in some $\pCFG$ and $\natfilt$ is (the smallest) sigma-algebra 
such that all the functions $\cfg{\sigma}{i}$, where $i\in \Nset_0$ and 
$\sigma$ is a scheduler, are $\natfilt$-measurable. In such a setting we can 
also consider sub-sigma-algebras $\natfilt_i$, $i\in \Nset_0$, of 
$\natfilt$, where $\natfilt_i$ is the smallest sub-sigma-algebra of 
$\natfilt$ such that all the functions $\cfg{\sigma}{j}$, $0\leq j \leq 
i$, are $\natfilt_i$-measurable. 
Intuitively, each set $A\in\natfilt_i$ consists of runs whose first $i$ steps 
satisfy some property.
%property, and the probability space $(\Omega,\natfilt_i,\probm^\sigma)$ 
%allows us to reason about probabilities of certain events happening in the 
%first 
%$i$ steps of program execution. 
%(note that since $\natfilt_i \subseteq 
%\natfilt$, we can view $\probm^\sigma$ also as a function of type 
%$\natfilt_i\rightarrow [0,1]$). 
Then, for each $A\in \natfilt_i$, the 
value $\E[\E[X|\natfilt_i]\cdot 1_A]$ represents the expected value of 
$X(\run)$ for the randomly generated run $\run$ provided that we restrict to 
runs whose
prefix of length $i$ satisfies the property given by $A$. 
The sequence $\natfilt_0,\natfilt_1,\natfilt_2,\dots$ forms a 
filtration of $\natfilt$, which we call a \emph{canonical filtration}.
\end{remark}

@@

While lexicographic ranking supermartingales provide a multidimensional extension of ranking supermartingales, 
there is an important difference even in the single dimension, which we explain below.
The strict 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{HolgerPOPL,CFNH16:prob-termination}. There is one significant difference: in these works there is an additional \emph{integrability} condition imposed on the one-dimensional process $\{X_i\}_{i=0}^{\infty}$, which requires that for each $i\geq 0$ it holds $\E[|X_i|]<\infty$ (or equivalently $\E[X_i]<\infty$, as the process is required to be non-negative). We do not impose this condition, which simplifies possible application of LexRSMs to programs with non-linear arithmetic, where, as already shown in~\cite{HolgerPOPL}, integrability of program variables is not guaranteed.
%We do not impose this condition, and indeed, it is possible to define a 3-dimensional LexRSM $\{\vec{X}_{i}\}_{i=0}^{\infty}$ such that $\E[X[3]_i]=\infty$ for some $i$ (example in \AppendixMaterial). 
The reason why integrability condition can be dropped is that it is only needed in the previous works to ensure that the conditional expectations exist and are well-defined. However, the existence of conditional expectations is also guaranteed for random variables that are real-valued non-negative, see Proposition~\ref{prop:conditional-exp-existence}. 
%This is exactly the case in both the original 1-dimensional ranking supermartingales and our generalization to LexRSMs.

### 2017-07-07 paper/synthesis.tex
Several small changes

The main idea is to iteratively synthesize 1-dimensional linear expression map that $1$-rank a subset of generalized transitions. These maps form the individual components of the sought-after $1$-LinLexRSM map. In each iteration, we start with a set $U$ of the yet-unranked generalized transitions. We seek a 1-dimensional LEM which ranks the maximal number of elements if $U$, and is unaffected by the remaining elements of $U$ (here by ranking a generalized transition $\tilde\tau$ we mean ranking it from each configuration $(\loc,\vec{x})$, where $\loc$ is the source location of $\tilde\tau$ and $\vec{x}\in\inv(\loc)$). If no 1-dimensional LEM that would rank at least one element in $U$ exists, then there is no LinLexRSM map for the program. Otherwise, we remove the newly ranked elements from $U$ and continue into the next iteration, until $U$ becomes empty. The process is summarized in Algorithm~\ref{algo:linlexrsm}.

@@

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
can be proved by generalization of arguments presented 
in~\cite{ADFG10:lexicographic}. 
\begin{theorem}
\label{thm:algo}
\label{THM:ALGO}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\overrightarrow{\lem}=(\lem_1,\dots,\lem_d)$, then 
	$\overrightarrow{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. If guards of all conditionals and loops in $\APP{}$ 
	are linear assertion, then the
	algorithm runs in time polynomial in size of 
	$\program$ and $\inv$.
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the 
$i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus 
U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since 
$U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 
1-ranked by some component of $\overrightarrow{\lem}$. Non-negativity of each 
$\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that 
each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this 
follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the 
fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.

>>>>

The main idea is to iteratively synthesize 1-dimensional {\em linear expression map (LEM)} that $1$-rank a subset of generalized transitions. These maps form the individual components of the sought-after $1$-LinLexRSM map. In each iteration, we start with a set $U$ of the yet-unranked generalized transitions. We seek a 1-dimensional LEM which ranks the maximal number of elements if $U$, and is unaffected by the remaining elements of $U$ (here by ranking a generalized transition $\tilde\tau$ we mean ranking it from each configuration $(\loc,\vec{x})$, where $\loc$ is the source location of $\tilde\tau$ and $\vec{x}\in\inv(\loc)$). If no 1-dimensional LEM that would rank at least one element in $U$ exists, then there is no LinLexRSM map for the program. Otherwise, we remove the newly ranked elements from $U$ and continue into the next iteration, until $U$ becomes empty. The process is summarized in Algorithm~\ref{algo:linlexrsm}.

@@

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
can be proved by generalization of arguments presented 
in~\cite{ADFG10:lexicographic}. 
\begin{theorem}
\label{thm:algo}
\label{THM:ALGO}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\overrightarrow{\lem}=(\lem_1,\dots,\lem_d)$, then 
	$\overrightarrow{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. If guards of all conditionals and loops in $\program$
	are linear assertions, then the
	algorithm runs in time polynomial in size of 
	$\program$ and $\inv$.
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the 
$i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus 
U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since 
$U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 
1-ranked by some component of $\overrightarrow{\lem}$. Non-negativity of each 
$\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that 
each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this 
follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the 
fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.
